---
title: "Simulating the evolution of a single neutral mutation"
subtitle: "Assignment 1 - Computational Tools in Evolutionary Biology" 
author: "Lucía Martín Fernández"
output:
  html_document:
    df_print: paged
---

## 1. Parameters

Declaration of parameters as suggested in the Assignment explanation

```{r}
N <- c(100, 300, 1000, 3000)  # Population size
BN <- c(0.25, 0.5, 1)         # Bottleneck sizes
days <- 10000                 # Length of experiment
replicates <- 100000          # Number of replicates
```

## 2. Variables

Declaration and initialization of matrices to store results. 

```{r}

# Matrix to store mean fixation time (tfix) for each experiment

tfix_matrix <- matrix(0, length(N), length(BN))    # Initialize 0 matrix NxBN size
colnames(tfix_matrix) <- BN                        # Column names are each of the bN values
rownames(tfix_matrix) <- N                         # Row names are each of the N values

# Matrix to store mean fixation probability (pfix) for each experiment

pfix_matrix <- matrix(0, length(N), length(BN))    # Initialize 0 matrix NxBN size
colnames(pfix_matrix) <- BN                        # Column names are each of the bN values
rownames(pfix_matrix) <- N                         # Row names are each of the N values

```

## 3. Simulation

```{r}

# Loop for Population size
for (n in N){
  
  # Loop for Bottleneck size
  for (bn in BN){
    
    # Print conditions
    cat("\nPopulation size:",n)
    cat("\nBottleneck size:",bn)
    
    times_v<-c()    # Vector to store tfix 
    
    
    for (replica in 1:replicates){
      
      pop<-rep(0,n) # Initialization of the population as a N size vector,
                    # since we are only considering one gene, it doesn't have to be a matrix
 
      pop[1]<-1     # Introduction of one individual with the mutation in the population
      
      
      for (day in 1:days){
        
        # Choose at random the bn proportion of the population that will be able to reproduce
        offspring<-sample(1:n, n*bn, replace=TRUE)
        # Update population vector adding 1/b number of offspring
        pop<-c(rep(pop[offspring], 1/bn)) 
        
        
        if (all(pop == 0)){                  # Mutation is lost
          
          break
          
        }else if (all(pop == 1)){            # Mutation is fixed
          
            times_v<-append(times_v,day)     # Append tfix
            
            break
        }
      
      }
    }
    
    pfix<-length(times_v)/replicates    # pfix = the number of times the fixation occurs / number of replicates 
    tfix<-mean(times_v)                 # tfix = mean of all tfix replicates
    
    # Print output
    cat("\nTfix:",tfix,"Pfix:",pfix,"\n")
    cat("\n")
    
    # Add results to matrices
    tfix_matrix[as.character(n),as.character(bn)]<-tfix
    pfix_matrix[as.character(n),as.character(bn)]<-pfix
    
  }
}
```

## 4. Results representation

### 4.1. Pfix and Tfix matrices

```{r}
# Matrix output
pfix_matrix
tfix_matrix
```
### 4.2.Plots



#### 4.2.1. Fixation time vs Bottleneck size

```{r}

# Definition of properties of the plot
plot(0,type='n',main=expression("T"["fix"]* " vs Bottleneck"),  # Title
     xlab="Bottleneck size", ylab="Fixation time",              # Axis labels
     xlim=c(0,max(BN)), ylim=c(0,max(tfix_matrix)+200))         # Limits

for (i in 1:length(N)){   # Iteration over population size
  
  x = BN                  # Bottleneck sizes
  y = tfix_matrix[i,]     # Tfix for each Bottleneck size  
  points(x,y,col=i*1.5)   # Plot all points with different colors depending on population size
  
  linearmodel<-lm(y~x)                     # Adjusting points to a linear model 
  abline(linearmodel,col=i*1.5, lwd='2')   # Drawing line that adjust to this linear model
  
  # Getting equations of the model
  coeff<-coefficients(linearmodel)
  n = round(coeff[1],2)
  m = round(coeff[2],2)
  cat("\nPopulation ",N[i]," : tfix =",n,"+",m,"· BN")
  
  # Legend
  text(0, max(tfix_matrix)+200, labels="N =", col = "darkgrey")
  text(0, max(tfix_matrix)-200, labels=expression("R"^2*" ="), col = "darkgrey")
  text(0+0.1*i, max(tfix_matrix)+200,labels=N[i], col=i*1.5)
  text(0+0.1*i, max(tfix_matrix)-200,labels=round(summary(linearmodel)$r.square,4), col=i*1.5)
  
}
```

#### 4.2.2.Fixation time vs Population size

```{r}

# Definition of properties of the plot
plot(0,type='n',main = expression("T"["fix"]* " vs Population"),   # Title
     xlab="Population size", ylab="Fixation time",                 # Axis label
     xlim=c(0,max(N)), ylim=c(0,max(tfix_matrix)+200))             # Limits

for (i in 1:length(BN)){     # Iteration over bottleneck size
  
  x = N                      # Population sizes
  y = tfix_matrix[,i]        # Tfix for each Population size  
  points(x,y,col=i*1.5)      # Plot all points with different colors depending on bottleneck size
  
  linearmodel<-lm(y~x)                     # Adjusting points to a linear model 
  abline(linearmodel,col=i*1.5, lwd='2')   # Drawing line that adjust to this linear model
  
  # Getting equations of the model
  coeff<-coefficients(linearmodel)
  n = round(coeff[1],2)
  m = round(coeff[2],2)
  cat("\nBottleneck ",BN[i]," : tfix =",n,"+",m,"· N")
  
  # Legend
  text(0, max(tfix_matrix)+200, labels="BN =", col = "darkgrey")
  text(0, max(tfix_matrix)-200, labels=expression("R"^2*" ="), col = "darkgrey")
  text(0+350*i, max(tfix_matrix)+200,labels=BN[i], col=i*1.5)
  text(0+350*i, max(tfix_matrix)-200,labels=round(summary(linearmodel)$r.square,4), col=i*1.5)
  
}
```

#### 4.2.3. Fixation probability vs Bottleneck size

```{r}

# Definition of properties of the plot
plot(0,type='n',main=expression("P"["fix"]* " vs Bottleneck"),  # Title
     xlab="Bottleneck size", ylab="Fixation probability",       # Axis labels
     xlim=c(0,max(BN)), ylim=c(0,max(pfix_matrix)+0.002))       # Limits

for (i in 1:length(N)){    # Iteration over population size
  
  x = BN                   # Bottleneck sizes
  y = pfix_matrix[i,]      # Pfix for each Bottleneck size  
  points(x,y,col=i*1.5)    # Plot all points with different colors depending on population size
  
  linearmodel<-lm(y~x)                     # Adjusting points to a linear model 
  abline(linearmodel,col=i*1.5, lwd='2')   # Drawing line that adjust to this linear model
  
  # Getting equations of the model
  coeff<-coefficients(linearmodel)
  n = round(coeff[1],3)
  m = round(coeff[2],3)
  cat("\nPopulation ",N[i]," : pfix =",n,"+",m,"· BN")
  
  # Legend
  text(0, max(pfix_matrix)+0.002, labels="N =", col = "darkgrey")
  text(0+0.1*i, max(pfix_matrix)+0.002,labels=N[i], col=i*1.5)
  
}
```

#### 4.2.4. Fixation probability vs Population size

```{r}

# Definition of properties of the plot
plot(0,type='n', main = expression("P"["fix"]* " vs Population ~ Linear Model"), # Title
     xlab="Population size", ylab="Fixation probability",                        # Axis labels
     xlim=c(0,max(N)), ylim=c(0,max(pfix_matrix)+0.002))                         # Limits

for (i in 1:length(BN)){     # Iteration over bottleneck sizes
  
  x = N                      # Population sizes
  y = pfix_matrix[,i]        # Pfix for each population size  
  points(x,y,col=i*1.5)      # Plot all points with different colors depending on bottleneck size
  
  linearmodel<-lm(y~x)                     # Adjusting points to a linear model 
  abline(linearmodel,col=i*1.5, lwd='2')   # Drawing line that adjust to this linear model
  
  # Legend
  text(300, max(pfix_matrix)+0.002, labels="N =", col = "darkgrey")
  text(300, max(pfix_matrix)+0.001, labels=expression("R"^2*" ="), col = "darkgrey")
  text(300+300*i, max(pfix_matrix)+0.002,labels=BN[i], col=i*1.5)
  text(300+300*i, max(pfix_matrix)+0.001,labels=round(summary(linearmodel)$r.square,3), col=i*1.5)
  
}
```

We can see that this plot does not fit a linear regression model. We can try to fix it to a logistic regression

```{r}

# Definition of properties of the plot
plot(0,type='n', main = expression("P"["fix"]* " vs Population ~ Log-Linear Model"), # Title
     xlab="Population size", ylab="Fixation probability",                            # Axis labels
     xlim=c(0,max(N)), ylim=c(0,max(pfix_matrix)+0.002))                             # Limits

for (i in 1:length(BN)){     # Iteration over bottleneck sizes
  
  x = N                      # Population sizes
  y = pfix_matrix[,i]        # Pfix for each population size  
  points(x,y,col=i*1.5)      # Plot all points with different colors depending on bottleneck size
  
  logmodel<-lm(y~log(x))                     # Adjusting points to a log-linear model         
  nx = seq(from=0,to=3000, length.out=1000)  # Creating a new x vector to get more points
  ny = predict(logmodel,newdata=list(x=nx))  # Calculating new y values for the new x 
  matlines(nx,ny, lwd=2, col=i*1.5)          # Plotting curve that fits the new y according to the model
  
  # Legend
  text(0+300, max(pfix_matrix)+0.002, labels="N =", col = "darkgrey")
  text(0+300, max(pfix_matrix)+0.001, labels=expression("R"^2*" ="), col = "darkgrey")
  text(300+300*i, max(pfix_matrix)+0.002,labels=BN[i], col=i*1.5)
  text(300+300*i, max(pfix_matrix)+0.001,labels=round(summary(logmodel)$r.square,3), col=i*1.5)
  
}
```

This model fits our data better, but we can improve it. Let's try with a Log-Log regression model: 

```{r}

# Definition of properties of the plot
plot(0,type='n', main = expression("P"["fix"]* " vs Population ~ Log-Log Model"), # Title
     xlab="Population size", ylab="Fixation probability",                         # Axis labels
     xlim=c(0,max(N)), ylim=c(0,max(pfix_matrix)+0.002))                          # Limits

for (i in 1:length(BN)){     # Iteration over bottleneck sizes
  
  x = N                      # Population sizes
  y = pfix_matrix[,i]        # Pfix for each population size  
  points(x,y,col=i*1.5)      # Plot all points with different colors depending on bottleneck size
  
  loglogmodel<-lm(log(y)~log(x))                  # Adjusting points to a log-log model         
  nx = seq(from=0,to=3000, length.out=1000)       # Creating a new x vector to get more points
  logy = predict(loglogmodel,newdata=list(x=nx))  # Calculating new log(y) values for the new x 
  ny = exp(logy)                                  # New y are the inverse of log(y), in other words exp(log(y))
  matlines(nx,ny, lwd=2, col=i*1.5)               # Plotting curve that fits the new y according to the model
  
  # Getting equations of the model
  coeff<-coefficients(loglogmodel)
  n = round(coeff[1],2)
  m = round(coeff[2],2)
  cat("\nBottleneck ",BN[i]," : log(tfix) =",n,"+",m,"· log(N)")
  
  # Legend
  text(0+300, max(pfix_matrix)+0.002, labels="N =", col = "darkgrey")
  text(0+300, max(pfix_matrix)+0.001, labels=expression("R"^2*" ="), col = "darkgrey")
  text(300+300*i, max(pfix_matrix)+0.002,labels=BN[i], col=i*1.5)
  text(300+300*i, max(pfix_matrix)+0.001,labels=round(summary(loglogmodel)$r.square,3), col=i*1.5)
  
}

```

The Log-Log Model fits the data perfectly
